---
title: "GBif data cleaning"
author: "Julian Avila-Jimenez"
date: "20/8/2020"
output:
  html_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = T, message = F, warning = F)
library(tidyverse)
```


Aplication exercise for search and cleaning occurrences from [GBif](https://www.gbif.org/) 

### Get the data

Search for a species of interest [*Puma concolor*](https://es.wikipedia.org/wiki/Puma_concolor)

```{r}
library(rgbif)
#obtain data from GBIF via rgbif
dat <- occ_search(scientificName = "Puma concolor", limit = 5000,
                  return = "data", hasCoordinate = T)

#names(dat$data) #a lot of columns

#select columns of interest
dat <- dat$data %>%
  dplyr::select(species, decimalLongitude, decimalLatitude, countryCode,
                individualCount,gbifID, family, taxonRank,
                coordinateUncertaintyInMeters, year,
                basisOfRecord, institutionCode, datasetName)
names(dat)
dim(dat)
```

### Record cleaning

First let's look the data

-remove records without coordinates

```{r}
dat <- dat %>%
  filter(!is.na(decimalLongitude))%>%
  filter(!is.na(decimalLatitude))

dim(dat)
```

-Visualize the data on a map
```{r}
library(ggplot2)

#plot data to get an overview
wm <- borders("world", colour="gray70", fill="gray80")
ggplot()+ 
  coord_fixed()+ 
  wm +
  geom_point(data = dat, 
             aes(x = decimalLongitude, y = decimalLatitude),
             colour = "red", size = 0.9)+
  theme_bw()+ 
  labs(title = paste("Spatial distribution of records",dat$species))
```

Then we will mark all suspicious records

```{r}
#Use CoordinateCleaner to automatically flag problematic records


library(countrycode)
#convert country code from ISO2c to ISO3c
dat$countryCode <-  countrycode(dat$countryCode, 
                                origin =  'iso2c', 
                                destination = 'iso3c')


library(CoordinateCleaner)
#flag problems
dat <- data.frame(dat)
flags <- clean_coordinates(x = dat, 
                           lon = "decimalLongitude", lat = "decimalLatitude",
                           countries = "countryCode", 
                           species = "species",
                           tests = c("capitals", "centroids", "equal",
                                     "gbif", "institutions",
                                     "zeros")) # most test are on by default

summary(flags)

plot(flags, lon = "decimalLongitude", lat = "decimalLatitude")

#Exclude problematic records
dat_cl <- dat[flags$.summary,]
dim(dat_cl)
#The flagged records
dat_fl <- dat[!flags$.summary,]
dim(dat_fl)
```

### Improving data quality using GBIF meta-data
  
-Remove records with low coordinate precision
```{r}
hist(dat_cl$coordinateUncertaintyInMeters / 1000, breaks = 20)

dat_cl <- dat_cl %>% 
  dplyr::filter(coordinateUncertaintyInMeters / 1000 <= 100 | is.na(coordinateUncertaintyInMeters))
```

### Remove unsuitable data sources

```{r}

table(dat$basisOfRecord)
dat_cl <- filter(dat_cl, basisOfRecord == "HUMAN_OBSERVATION")

table(dat_cl$basisOfRecord)

#suspicious individual counts removal 
#Individual count
table(dat_cl$individualCount)

dat_cl <- dat_cl%>%
  filter(individualCount > 0 | is.na(individualCount))%>%
  filter(individualCount < 99 | is.na(individualCount)) # high counts are not a problem

```

### Check for old records

```{r}
#remove very old records

#Age of records
table(dat_cl$year)

dat_cl %>% 
  group_by(year) %>% 
  summarise(records=n()) %>% 
  ggplot(aes(year, records))+
  geom_line(color="steelblue")+
  labs(title=paste("Number of records of:",dat_cl$species, "by year"),
       y="",x="")+
  theme_bw()
```

### Final publication map

```{r}

library("rnaturalearth")
library(sf)
library(ggspatial)
world <- ne_countries(scale = "medium", returnclass = "sf")

coords<-dat_cl %>% 
  dplyr::select(species,decimalLongitude, decimalLatitude ) %>%
  rename(Longitude=decimalLongitude, Latitude=decimalLatitude)

coords_sf<- coords %>% 
  st_as_sf(coords=(2:3))

bbox_new <- st_bbox(coords_sf)


ggplot()+
  geom_sf(data = world, fill= NA) +
  coord_sf(xlim=c(bbox_new[1],bbox_new[3]), 
           ylim=c(bbox_new[2],bbox_new[4]))+
    geom_point(data = dat_cl, aes(x = decimalLongitude, y = decimalLatitude),
             colour = "cyan4", size = 1.2, alpha=0.6)+
  annotation_scale(location = "br", width_hint = 0.1,line_width = 0.5) +
  annotation_north_arrow(location = "tr", which_north = "true", 
                         pad_x = unit(1.5, "cm"), 
                         pad_y = unit(0.5, "cm"),
                         height = unit(1, "cm"),
                         width = unit(1, "cm"), # 0.2 # 0.3
                         style = north_arrow_fancy_orienteering)+
  labs(x = "", y = "",
       title =paste("Record distribution of:", dat_cl$species),
       caption = paste("Source: GBif\nAuthor: Julián Avila- Jiménez\nDate:",
                       format(Sys.time(), '%d %B, %Y')))+
  theme(plot.title=element_text(size=19),
        plot.caption = element_text(size = 7, color="grey60"),
        panel.background= element_rect(fill = "grey96"))

```

### Leaflet visualization

```{r}
library(leaflet)
leaflet (dat_cl) %>%
  addProviderTiles("Stamen.Terrain") %>% 
  addCircleMarkers(lng=~decimalLongitude, 
                   lat=~decimalLatitude,
                   fillColor = "red",
                   color = "red",
                   radius = 5,
                   opacity = 0.7,
                   weight = 1) 

```

